### 2024-07-07 cognitive-bias.qmd
20240706 - characteristics of science vs pseudoscience

Statistical analysis—and science more generally—is a process to the pursuit of knowledge.
An *epistemology* is an approach to knowledge.
Science is perhaps the best approach (epistemology) that society has to approximate truth.
Unlike other approaches to knowledge, science relies on empirical evidence and does not give undue weight to anecdotal evidence, intuition, tradition, or authority.
Nevertheless, statistical analysis is not purely objective and is not a panacea.
Science is a human enterprise—it is performed by humans each of whom has their own biases.
For instance, cognitive biases can influence how people interpret statistics.
As a result, the findings from any given study may be incorrect.
Thus, it would be imprudent to make decisions based solely on the results of one study.
That is why we wait for findings to be independently replicated by different groups of researchers using different methods.

>>>>

Statistical analysis—and science more generally—is a process to the pursuit of knowledge.
An *epistemology* is an approach to knowledge.
Science is perhaps the best approach (epistemology) that society has to approximate truth.
Unlike other approaches to knowledge, science relies on empirical evidence and does not give undue weight to anecdotal evidence, intuition, tradition, or authority.

Per @PetersenPrinciplesPsychAssessment, here are the characteristics of science that distinguish it from pseudoscience:

> 1. Risky hypotheses are posed that are falsifiable.
> The hypotheses can be shown to be wrong.
> 1. Findings can be replicated independently by different research groups and different methods.
> Evidence converges across studies and methods.
> 1. Potential alternative explanations for findings are specified and examined empirically (with data).
> 1. Steps are taken to guard against the undue influence of personal beliefs and biases.
> 1. The strength of claims reflects the strength of evidence.
> Findings and the ability to make judgments or predictions are not overstated.
> For instance, it is important to present the degree of uncertainty from assessments with error bars or confidence intervals.
> 1. Scientifically supported measurement strategies are used based on their psychometrics, including [reliability](#sec-reliability) and [validity](#sec-validity).

Nevertheless, statistical analysis is not purely objective and is not a panacea.
Science is a human enterprise—it is performed by humans each of whom has their own biases.
For instance, cognitive biases can influence how people interpret statistics.
As a result, the findings from any given study may be incorrect.
Thus, it would be imprudent to make decisions based solely on the results of one study.
That is why we wait for findings to be independently replicated by different groups of researchers using different methods.
